---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
###Initialization
```{r}
library(plyr)
library(lattice)
```

## Loading and preprocessing the data
Load the dataset and have a look at its structure, and data types:
```{r}
activity <- read.csv("activity.csv")
str(activity)
```
Ok, so the "date" columns is currently a factor. Let's do something about that:
```{r}
activity[,2] <- as.Date(activity$date)
```
Let's have a look at the structure again:
```{r}
str(activity)
```


## What is mean total number of steps taken per day?
Sweet, now that the dataset is in order, let's have a look at the characteristics of the data in terms of mean and median:
```{r}
steps <- with(activity,tapply(steps,date,sum, na.rm=T))
steps_mean <- round(mean(steps),1)
steps_median <- median(steps)
print(paste0("Mean steps per day is: ", steps_mean, ", and median: ", steps_median))
```
I see that this person has a better score than me, as I'm mostly just sitting on my arse in my office. Well well, always room for improvement.
  
Let's do a histogram:
```{r}
steps_withNAs <- with(activity,tapply(steps,date,sum, na.rm=F))
hist(steps_withNAs, xlab = "Steps per day", main = "Histogram of number of steps per day (with NA's)")
```


## What is the average daily activity pattern?
A time series plot over average number of steps (x-axis) take per interval accross all days (y-axis):
```{r}
mean_steps_per_interval <-with(activity,tapply(steps,interval,mean,na.rm=T))
intervals<-unique(activity$interval)
df_avgsteps_interval <-data.frame(cbind(mean_steps_per_interval,intervals))
plot(df_avgsteps_interval$intervals,
    df_avgsteps_interval$mean_steps_per_interval,
    type = "l",
    xlab = "Intervals",
    ylab = "Average Steps",
    main = "Average Steps per Interval")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
index<-which.max(df_avgsteps_interval$mean_steps_per_interval)
print(df_avgsteps_interval[index,2])
```

As we saw earlier, the "steps" column has a lot of NA's. How many?
```{r}
sum(is.na(activity$steps))
```


## Imputing missing values
Next task is to replace the NA's in steps to later see the consequences of not taking these into account. First, let's create a copy of the activity data set where the NA's are replaced with the average number of steps per day.
```{r}
index<-which(is.na(activity$steps))
l<-length(index)
steps_avg<-with(activity,tapply(steps,date,mean,na.rm=T))
na<-mean(steps_avg,na.rm = T)

##Create a copy of activity where NA's are replaced
activity_noNA <- activity
for (i in 1:l) {
        activity_noNA[index[i],1]<-na
}
```
  
Let's compare the old and new version of the activity dataset:
```{r}
head(activity)
head(activity_noNA)
```
  
And then move on to compare histograms over first and new version:
```{r}
steps_noNAs <- with(activity_noNA,tapply(steps,date,sum))
hist(steps_noNAs, ylim = c(0,40), xlab = "Steps per day", main = "Histogram of number of steps/day (NA's replaced with an average)")
steps_withNAs <- with(activity,tapply(steps,date,sum, na.rm=F))
hist(steps_withNAs,ylim = c(0,40), xlab = "Steps per day", main = "Histogram of number of steps/day (with NA's)")
```
  
Whatabout changes in mean and median between the two datasets?
```{r}
steps_noNAs_mean <- round(mean(steps_noNAs),1)
steps_noNAs_median <- round(median(steps_noNAs),1)
print(paste0("Mean steps when NA's still there: ", steps_mean, ", and without NA's: ", steps_noNAs_mean))
print(paste0("Median steps when NA's still there: ", steps_median, ", and without NA's: ", steps_noNAs_median))
```


## Are there differences in activity patterns between weekdays and weekends?
On a different note, is there a difference in pattern between weekdays and weekends?
(using the dataset where NA's are replaced with an average)
  
Let's first categorize each day as a weekday/weekend, and add that category as a factor:
```{r}
activity_noNA$DayCategory <- ifelse(weekdays(activity_noNA$date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

```
  
Now, let's look at the difference in pattern:
```{r}
## Summarize data by interval and type of day
df_avgsteps_interval_2 <- ddply(activity_noNA, .(interval, DayCategory), summarize, Avg = mean(steps))
head(df_avgsteps_interval_2)

##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, 
       data=df_avgsteps_interval_2, 
       type="l", 
       layout = c(1,2),
       main="Average Steps per Interval based on type of day",
       ylab="Average Number of Steps",
       xlab="Interval")

```
